---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey (ACS) data"
author: "RYSKA"
date: 2026-09-10
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(tidycensus)
```



---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


**In this analysis, I look at median home value across counties in Illinois state, and ask which communities we measure most and least reliably.**

---

## Choosing what to measure


```{r}
my_state <- "IL"        # two-letter abbreviation, in quotes
my_variable <- "B25077_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```



**There are 102 rows in the data that match the number of counties in the Illinois**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```



Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```



**In this data, Hardin County has the largest margin of error relative to its estimate with 34.09%**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```



**There are 49 counties with high confidence in reliability (moe percentage < 5%), 42 counties with moderate confidance in reliablity (5% < moe percentage < 10%), and 11 counties with low confidence reliability (moe percentage > 10%)**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```



```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```



**Average population in high confidence counties is the largest (240,766 people) compared to the other groups. On the other hand, the low confidence counties have the smallest average population with 4,332 people. Small counties tend to have low confidence reliability, while large counties tend to have high confidence reliability**

---

## What this means for policy


**Census data is commonly used by policymakers as a source to develop programs. However, not all policies are perfectly suited to every area. There are many reasons for this. From an analytical perspective, understanding data sources reliability is important for identifying the limitations of an analysis and knowing how to address them, so we could get a better understanding of places to make a better decision. The ACS uses sampling methods to get estimation number. Therefore, it has a margin of error (MOE), which represents the range of uncertainty of an estimate.** 

**Looking at median home value data from the ACS 5-year estimates for Illinois, the MOE percentage varies from 0.6% to 34.1%. With the largest MOE percentage (34.1%), Hardin County is the area about which we have the least certainty. There is a pattern in which low confidence reliability counties has the least average population, when high confidence reliability counties has the highest. Because the ACS uses sampling methods, smaller places yield smaller samples, which tend to result in lower reliability.**

**Understanding data reliability is important. If not, we may treat data from all counties as equally certain and apply policies without considering that data from smaller counties may be less representative of actual conditions. In this case, relying solely on median home value data in Illinois without considering its reliability could lead to housing assistance program application that are not well suited to Hardin County or other lower confidence counties, and smaller communities may continue to be disadvantaged.**

---


